How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in \textit{Emys marmorata} (Testudinoidea, Emydidae).

### 2013-08-07 documents/turtle_paper.Rnw
most of ken comments. found early error in mung file. need to fix.

\documentclass[12pt,letterpaper]{article}
\usepackage{amsmath, amsthm}
\usepackage{graphicx, microtype, hyperref, authblk}
\usepackage{rotating, longtable, caption, subcaption, multirow}
\usepackage[sort&compress]{natbib}
\usepackage{fullpage}

% for american naturalist
%\usepackage{lineno}

@@

\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation.}
\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
\author[2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
\affil[1]{Committee on Evolutionary Biology, University of Chicago}
\affil[2]{Department of Geology, Field Museum of Natural History}
\affil[3]{Department of Geological Sciences, California State University -- Fullerton}

@@

%\linenumbers
%\modulolinenumbers[2]

@@

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection{Background and system}
Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004}. These methods are comparatively simple and straight forward ways of understanding the differences in morphology between classes. Also, they are very visual methods which aides in interpretation and presentation of information. Neural network models have also been introduced and applied frequently in the context of automated taxon identification along with more general applications \citep{MacLeod2007}. Here, we used multiple alternative machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses. These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on. Additionally, we investigate variation in continuous traits, and do not search for discrete differences between each class, instead focusing on suites of traits together.

% this paragraph kind of sucks
The two major classes of machine learning methods, unsupervised and supervised, are essentially extensions of known statistical methods. Unsupervised learning methods are analogous to clustering and density estimation methods, while supervised learning methods are analogous classification and regression models. In both cases, many of these methods are not fit via maximum likelihood and are supplemented by randomization, sorting, and partitioning algorithms along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled. The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models. Additionally, instead of pure classification accuracy, here we use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified (see Methods).

@@

Previous work on morphological differentiation between different populations of \textit{E. marmorata} with no alternative subspecific assignments \citep{Seeliger1945,Holland1992}. \citet{Holland1992} used comparisons of linear measures of the carapace of \textit{E. marmorata} from different populations over the entire range of the species. This study was interested in the relative effect of distance versus barriers had in terms of fostering morphological differentiation in \textit{E. marmorata}. Analyses were performed to determine how different, morphologically, different populations in three different regions of the species range. Additionally, the degree of sexual dimorphism present in \textit{E. marmorata}. The study concluded that distance was a poor indicator of morphological differentiation as opposed to barriers, such as different drainage basins, are probably more important barriers to reproduction. This conclusion was later echoed by \citet{Spinks2005} via molecular phylogenetic analysis. Additionally, \citet{Holland1992} found that with increasing amount of barriers and distance, morphological differentiation was observable though the underlying variation required many variables obtain indicating the very fine degree of morphological differentiation between putatively distinct populations. \citet{Holland1992} concluded that \textit{E. marmorata} is best classified as three distinct species as opposed to subspecies: a northern species, southern species, and Columbia basin species. This classification is similar to \citet{Seeliger1945}, except elevated to the species as opposed to subspecific level.

@@

\section{Materials and Methods}
\subsection{Specimens}
We collected landmark-based morphometric data from \Sexpr{nrow(turtle.adult)} adult \textit{E. marmorata} museum specimens. Specimen classification was based on known specimen geographic information which was recorded from museum collection information. When precise latitude and longitude information was not available it was estimated from whatever locality information was present. Because the specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all specimen classifications were based solely on this geographic information and not from explicit assignment in previous studies. Because the exact barriers between different biogeographic regions are unknown and unclear, two assignments for both the morphologically and molecularly based hypotheses were used. Each morphologically based hypothesis had three classes, while each molecular-based had four classes. In total, each specimen was given four different classifications.

\subsection{Geometric morphometrics}
% landmarks
Following \citet{Angielczyk2011}, 19 landmarks were digitized using TpsDig 2.04 \citep{Rohlf2005}. These landmarks were chosen to maximize the description of general plastral variation. 17 of these landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron. 12 of these landmarks were chosen to be symmetrical across the axis of symmetry and, in order to prevent degrees of freedom and other concerns \citep{Klingenberg2007}, prior to analysis these landmarks were reflected across the axis of symmetry (i.e. midline) and the average position of each symmetrical pair was used. In cases where damage or incompleteness prevented symmetric landmarks from being determined, only the single member of the pair was used. Analysis was conducted on the resulting ``half'' plastra.
% GP superimposition and PCA
Plastral landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components (PC) of shape were calculated. This was done using the \texttt{shapes} package for R \citep{2013,Dryden2013}.

@@

%   PAM
The \(\rho\) dissimilarity matrix was divisively clustered using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
%   gap statistic
Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of clusters varied between one and \Sexpr{nrow(tadult.gap$Tab)}. Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}. The gap statistic is defined
\[Gap_{n}(k) = E^{*}_{n}[\log(W_{k})] - \log(W_{k})\] 
where \(W_{k}\) is
\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and there respective mediods (\(C\)) for all clusters \(r\). This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution. In this case, the reference distribution was estimated from \Sexpr{tadult.gap$B} resamples of the dataset taking into account the original structure of the data.
%  R implementation
This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all \Sexpr{nrow(turtle.adult)} observations.

\subsubsection{Supervised learning}
% training testing split
The total dataset of \Sexpr{nrow(turtle.adult)} observations was split into training and testing datasets. The training dataset represented 75\% of the total dataset, split proportionally by class, and was used for model fitting. The testing dataset represented the remaining 25\% of the total dataset and  was used after model fitting to estimate the effectiveness of each classification hypothesis and generalizability of the supervised learning models (i.e. performance in the wild).

@@

%     recursive feature selection
Random forest models are not fit using maximum likelihood so AICc based model selection was not possible. Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model. Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features were included in the initial model, there importance ranked, and the AUC of the model calculated. The lowest ranked feature was then removed, and the AUC of the model recalculated. This was repeated until only one feature, remained. Similar to the multinomial logistic regression models described above, the maximum number of PCs that could have been included in the model was \Sexpr{floor(max.ad)}. After each PC was removed , 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model. Random forest model parameters were estimated from 1000 subtrees. Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, this means that the exact PCs included in each model do not correspond to the PCs in each of the 10 multinomial logistic regressions models.

@@

\section{Results}
\subsection{Geometric morphometrics}
The results of the PCA of the total dataset of \textit{E. marmorata} pastral landmarks configurations demonstrates no clear or obvious groupings (Fig. \ref{fig:pca}). The first three PCs, which represent \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the total variation, are a cloud of points with no structure. Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig. \ref{fig:pca}). However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark \citep{Zelditch2004} and this observation should be considered cursory at best.
The first PC appears to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig. \ref{fig:pc_var}). Variation along the second PC appears to be characterized by variation along the midline, especially in terms of the ``straightness'' of the individual landmarks along it.

@@

% PAM results
Comparison of gap statistic values for the range of PAM solutions indicates that the optimal number of clusters is one (Fig. \ref{fig:gap}). The next best clustering solution had only two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed (Fig. \ref{fig:gap_map}). Importantly, these clusters do not conform to the nothern and souther groups from the nuclear DNA hypothesis \citep{Spinks2010}.

Unfortunately, it was not possible to obtain enough detailed information on the sex of each \textit{E. marmorata} specimen, thus it is difficult at best to determine if this clustering solution corresponds to sexual dimorphism between observations. Male emydid turtles are known to have a plastral concavity which may influence landmark position along the midline. However, the plastral concavity of \textit{E. mamorata} males is considered less pronounced than in other emydid turtles. While we cannot completely rule out sexual dimorphism as the root cause of this observation, because of the very small degree of known sexual dimorphism in \textit{E. marmorata} we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.

@@

The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that one of the molecular classification hypotheses was the best overall classification scheme (Fig. \ref{fig:gen_res}). For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:mm_test} and \ref{tab:rf_test}). Remarkably, the best classification hypothesis was identical based on both the multinomial logistic regression and random forest models.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width = \textwidth]{figure/pca_imp}
  \caption{Pairs plot of the first three most important variables of the optimal random forest model of turtle plastral shape. The variables descend in importance from the upper left to the lower right. The observations are colored as in figure \ref{fig:gen_map}.  The bottom row are histograms of classification occurrences along the PCs.}
  \label{fig:imp_pc}
\end{figure}
% need to add in variance percentages on the different axes.

@@

%RELATIVE RISK FROM THE MULTINOMIAL LOGISTIC REGRESSION MODELS
The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that while an individual PC might not be sufficient to differentiate any individual class from the northern group, given multiple features the odds of determining the correct classification increase (Fig. \ref{fig:rel_risk}). Changes along the second most important axes contribute very little to increasing the odds of classification. This is observable from the class histograms of PC 8 (Fig. \ref{fig:imp_var}). Changes along the first and third most important axes contribute more obviously to increasing the odds of correctly identifying the class of an observation, a result that is observable in both the relative risk (Fig. \ref{fig:rel_risk}) and the different class histograms of the PCs (Fig. \ref{fig:imp_var}).

@@

%   unsupervised learning
%     model based clustering
%     nonparametric bayesian approach
%     anything about semisupervised clustering? probably not
Ultimately, it would be useful to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa. The only unsupervised method employed in this study, PAM, is rather simple and not model based. A more useful approach would be to employ various model based clustering approaches \citep{Fraley2002,Zhong2003}. In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors \citep{Fraley2002}, in order to asses the best clustering solution.

% future directions
%   potentially useful for phylogenetic comparative method stuff
%   conservation utility
%   fossil diversity
%     current approach favors having an extremely explicit classification hypothesis
%     as indicated about, nonparametric bayes and other model based clustering can help

@@

\section{Acknowledgements}
PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study. MUSEUMS. GRANTING AGENCIES.

@@

\begin{table}
  \centering
  \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
    \cline{3-4} 
    & & \multicolumn{2}{ c |}{Predicted class} \\ \cline{3-4}
    & & 1 & 0 \\ \hline
    \multicolumn{1}{| c |}{\multirow{2}{*}{Actual class}}
    & 1 & TRUE \newline POSITIVE & FALSE \newline NEGATIVE \\ \cline{2-4}
    \multicolumn{1}{| c |}{} & 0 & FALSE \newline POSITIVE & TRUE \newline NEGATIVE \\
    \hline
  \end{tabular}
  \caption{Example confusion matrix. The columns correspond to the predicted class of an observation, while the rows correspond to the actual class of that observation. Depending on the type match between the prediction and reality, four different outcomes are possible: true positive (TP), false negative (FN), false positive (FP), and true negative (TN). These four quantities are used for calculating all confusion matrix statistics. Each of these values is an integer and the sum of the number of occurrences of that event during classification.}
  \label{tab:conf}
\end{table}

>>>>

\documentclass[12pt,letterpaper]{article}
\usepackage{amsmath, amsthm}
\usepackage{graphicx, morefloats, microtype, hyperref, authblk}
\usepackage{rotating, longtable, caption, subcaption, multirow}
\usepackage[sort&compress]{natbib}
\usepackage{fullpage}

% for american naturalist
\usepackage{lineno}

@@

\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in \textit{Emys marmorata} (Testudinoidea, Emydidae).}
\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
\author[1,2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
\affil[1]{Committee on Evolutionary Biology, University of Chicago}
\affil[2]{Integrative Research Center, Field Museum of Natural History}
\affil[3]{Department of Geological Sciences, California State University -- Fullerton}

@@

\linenumbers
\modulolinenumbers[2]

@@

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection{Background and system}
Differences in morphological variation between different classes has previously been analyzed using methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004}. These methods are comparatively simple and straight forward ways of understanding the differences in morphology between classes. Also, they are very visual methods which aides in interpretation and presentation of information. Neural network models have also been introduced and applied frequently in the context of automated taxon identification along with more general applications \citep{MacLeod2007}. Here, we used multiple alternative machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses. These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on. Additionally, we investigate variation in continuous traits, and do not search for discrete differences between each class, instead focusing on multivariate quantification of shape.

% this paragraph kind of sucks
The two major classes of machine learning methods, unsupervised and supervised, are essentially extensions of known statistical methods \citep{Hastie2009}. Unsupervised learning methods are analogous to clustering and density estimation methods \citep{Kaufman1990}, while supervised learning methods are analogous classification and regression models \citep{Breiman1984}. In both cases, many of these methods are not fit via maximum likelihood and are supplemented by randomization, sorting, and partitioning algorithms along with the maximization or minimization of summary statistics in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}. The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models. 
% Applications of these methods outside of this study
% epidemeology, computer science, marketing
Additionally, instead of pure classification accuracy, here we use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified (see Methods).

@@

Previous work on morphological differentiation between different populations of \textit{E. marmorata} with no alternative subspecific assignments \citep{Seeliger1945,Holland1992}. 
%
\citet{Holland1992} used comparisons of linear measures of the carapace of \textit{E. marmorata} from different populations over the entire range of the species. This study was interested in the relative effect of distance versus barriers had in terms of fostering morphological differentiation in \textit{E. marmorata}. Analyses were performed to determine how different, morphologically, different populations in three different regions of the species range. Additionally, sexual dimorphism present in \textit{E. marmorata} was only assesed in the context of size measures such as length of carapace and mass. In general, \citet{Holland1992} and others \citep{Lubcke2007,Germano2008} found that males of \textit{E. marmorata} are larger than females in terms of size. However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a} and \citet{Dryden1998a}, was not assessed \citep{Holland1992,Lubcke2007,Germano2008}.

\citet{Holland1992} concluded that distance was a poor indicator of morphological differentiation as opposed to barriers, such as different drainage basins, are probably more important barriers to reproduction. This conclusion was later echoed by \citet{Spinks2005} via molecular phylogenetic analysis. Additionally, \citet{Holland1992} found that with increasing amount of barriers and distance, morphological differentiation was observable though the underlying variation required many variables obtain indicating the very fine degree of morphological differentiation between putatively distinct populations. \citet{Holland1992} concluded that \textit{E. marmorata} is best classified as three distinct species as opposed to subspecies: a northern species, southern species, and Columbia basin species. This classification is similar to \citet{Seeliger1945}, except elevated to the species as opposed to subspecific level.

@@

\section{Materials and Methods}
\subsection{Specimens}
We collected landmark-based morphometric data from \Sexpr{nrow(turtle.adult)} adult \textit{E. marmorata} museum specimens. These specimens include both newly sampled individuals and those sampled in previous studies of plastral shape variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}. 
Specimen classification was based on known specimen geographic information which was recorded from museum collection information. When precise latitude and longitude information was not available it was estimated from whatever locality information was present. Because the specimens used to define the subclades in \citet{Spinks2005} and \citet{Spinks2010} were not available for study, all specimen classifications were based solely on this geographic information and not from explicit assignment in previous studies. The classifications here were based on matching museum locality data with the geographic boundaries of the moelcularly-defined clades of \citet{Spinks2005} and \citet{Spinks2010}. Because the exact barriers between different biogeographic regions are unknown and unclear, two assignments for both the morphologically and molecularly based hypotheses were used. Each morphologically based hypothesis had three classes, while each molecular-based had four classes. In total, each specimen was given four different classifications.

\subsection{Geometric morphometrics}
% landmarks
Following previous work on plastral variation \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, 19 landmarks were digitized using TpsDig 2.04 \citep{Rohlf2005}. These landmarks were chosen to maximize the description of general plastral variation(Fig. \ref{fig:plastra}). 17 of these landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the platron. 12 of these landmarks were chosen to be symmetrical across the axis of symmetry and, in order to prevent degrees of freedom and other concerns \citep{Klingenberg2002}, prior to analysis these landmarks were reflected across the axis of symmetry (i.e. midline) and the average position of each symmetrical pair was used. In cases where damage or incompleteness prevented symmetric landmarks from being determined, only the single member of the pair was used. Analysis was conducted on the resulting ``half'' plastra.
% GP superimposition and PCA
Plastral landmark configurations were superimposed using generalized Procrustes analysis \citep{Dryden1998a} after which, the principal components (PC) of shape were calculated. This was done using the \texttt{shapes} package for R \citep{2013,Dryden2013}.

@@

%   PAM
The \(\rho\) dissimilarity matrix was divisively clustered using partitioning around mediods clustering (PAM), a method similar to \textit{k}-means clustering except that instead of minimizing the sum of squared Euclidean distances between observations and centroids, the sum of squared dissimilarities between observations and mediods is minimized \citep{Kaufman1990}.
%   gap statistic
Because the optimal number of clusters of shape configurations in the study was unknown, being possibly three, four, or some other value, clustering solutions were estimated with the number of clusters varied between one and \Sexpr{nrow(tadult.gap$Tab)}. Clustering solutions were compared using the gap statistic, which is a measure of goodness of clustering \citep{Tibshirani2001a}. The gap statistic is defined
\[Gap_{n}(k) = E^{*}_{n}[\log(W_{k})] - \log(W_{k})\] 
where \(W_{k}\) is
\[W_{k} = \sum^{k}_{r = 1}{\frac{1}{2n_{r}} (\sum_{i,i' \in C_{r}} d_{ii'})}\].
\(d_{ii'}\) is the dispersion of the clustering solution or the sum of the pairwise dissimilarities between observations in each cluster and their respective mediods (\(C\)) for all clusters \(r\). This value is averaged and compared to the expected dispersion (\(E^{*}_{n}\)) of a sample \(n\) from a reference distribution. In this case, the reference distribution was estimated from \Sexpr{tadult.gap$B} resamples of the dataset while maintaining the original dispersion of the data.
%  R implementation
This analysis was conducted using the \texttt{cluster} package for R \citep{Maechler2013} using all \Sexpr{nrow(turtle.adult)} observations.

\subsubsection{Supervised learning}
% training testing split
The total dataset of \Sexpr{nrow(turtle.adult)} observations was split into training and testing datasets. The training dataset represented 75\% of the total dataset, split proportionally by class, and was used for model fitting. The testing dataset represented the remaining 25\% of the total dataset and  was used after model fitting to estimate the effectiveness of each classification hypothesis and generalizability of the supervised learning models (i.e. performance in the wild). This split was chosen to allow for a large enough sample size for model fitting while also providing a large enough testing dataset to determine any systematic misclassifcations.

@@

%     recursive feature selection
Random forest models are not fit using maximum likelihood so AICc based model selection was not possible. Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model. Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features were included in the initial model, their importance ranked, and the AUC of the model calculated. The lowest ranked feature was then removed, and the AUC of the model recalculated. This was repeated until only one feature, remained. Similar to the multinomial logistic regression models described above, the maximum number of PCs that could have been included in the model was \Sexpr{floor(max.ad)}. After each PC was removed , 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model. Random forest model parameters were estimated from 1000 subtrees. Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, this means that the exact PCs included in each model do not correspond to the PCs in each of the 10 multinomial logistic regressions models.

@@

\section{Results}
\subsection{Geometric morphometrics}
The results of the PCA of the total dataset of \textit{E. marmorata} pastral landmarks configurations demonstrates no clear or obvious groupings (Fig. \ref{fig:pca}). The first three PCs, which represent \Sexpr{paste0(sum(turtle.fit$percent[1:3]))}\% of the total variation, are a cloud of points with no structure. Additionally, individual landmark variation is mostly circular around each landmark with some more elliptical variation observed along some midline landmarks and the most lateral landmark (Fig. \ref{fig:pca}). However, it is important to note that Procustes based superimposition attempts to evenly distribute variance around the mean landmark \citep{Zelditch2004} and this observation should be considered cursory at best.

The first PC appears to describe variation in lateral plastral shape, from a very pointed medial edge to a much more rounded and blunt edge (Fig. \ref{fig:pc_var}). Landmark 10 (Fig. \ref{fig:plastra}), which appears to be the most variable along this axis (Fig. \ref{fig:pca} and \ref{fig:pc_var}), is positioned on the bridge between the plastron and the carapace. Over ontogeny, this is an area that deepens dorsoventrally and when the plastron was projected into two dimensions this it created the effect of mediolateral movement.
Variation along the second PC appears to be characterized by variation along the midline, especially in terms of the ``straightness'' of the individual landmarks along it. This is most likely caused by the fact that plastral scutes do not always line up perfectly along the midline.

% size correlation with either of the first two PCs?

@@

% PAM results
Comparison of gap statistic values for the range of PAM solutions indicates that the optimal number of clusters is one (Fig. \ref{fig:gap}). The next best clustering solution had only two clusters, however there is no geographic structure to this classification scheme, with members of these clusters being seemingly randomly distributed (Fig. \ref{fig:gap_map}). Importantly, these clusters do not conform to the northern and southern groups from the nuclear DNA hypothesis \citep{Spinks2010}.

% some sex information. Analyze it.
Unfortunately, it was not possible to obtain enough detailed information on the sex of each \textit{E. marmorata} specimen, thus it is difficult at best to determine if this clustering solution corresponds to sexual dimorphism between observations. Male emydid turtles are known to have a plastral concavity which may influence landmark position along the midline. However, the plastral concavity of \textit{E. mamorata} males is considered less pronounced than in other emydid turtles. While we cannot completely rule out sexual dimorphism as the root cause of this observation, because of the very small degree of known sexual dimorphism in \textit{E. marmorata} we are less concerned with possible effects of sexual dimorphism for the later supervised learning methods.

@@

The results from the generalization of the selected supervised learning models, measured by the distributions of the bootstrapped AUC values of the testing dataset, show that the second molecular classification hypotheses was the best overall classification scheme (Fig. \ref{fig:gen_res}). For both methods, the distribution of bootstrapped AUC for the molecular hypothesis was significantly greater than all of the other classification schemes (Tables \ref{tab:mm_test} and \ref{tab:rf_test}). Remarkably, the best classification hypothesis was identical based on both the multinomial logistic regression and random forest models.

@@

@@

The relative risk values for classification from the multinomial logistic regression model, based on the three most important PCs, demonstrate that while an individual PC might not be sufficient to differentiate any individual class from the northern group, given multiple features the odds of determining the correct classification increase (Fig. \ref{fig:rel_risk}). Changes along the second most important axes contribute very little to increasing the odds of classification. This is observable from the class histograms of PC 8 (Fig. \ref{fig:imp_var}). Changes along the first and third most important axes contribute more obviously to increasing the odds of correctly identifying the class of an observation, a result that is observable in both the relative risk (Fig. \ref{fig:rel_risk}) and the different class histograms of the PCs (Fig. \ref{fig:imp_var}).

@@

%   unsupervised learning
Ultimately, it would be useful to not require such explicit classification hypotheses, especially when concerned about possible cryptic variation in extinct taxa. The only unsupervised method employed in this study, PAM, is rather simple and not model based. A more useful approach would be to employ various model based clustering approaches \citep{Fraley2002,Zhong2003}. In this manner, a series of candidate models can be compared via model comparison methods, such as AIC or Bayes factors \citep{Fraley2002}, in order to asses the best clustering solution. % beat complaints from reviewers
Here we focused on the results and utility of supervised methods because they are both more powerful and hypothesis driven \citep{Hastie2009}. Because there are two alternative classification schemes for \textit{E. marmorata}, it was most appropriate to compare these two hypotheses and estimate which one most accurately reflected the variation. Future work would be to explore and derive unsupervised methods which corroborate these results.

@@

\section{Acknowledgements}
PDS would like to thank David Bapst, Michael Foote, Benjamin Frable, and Dallas Krentzel for useful discussion which enhanced the quality of this study. For access to emydine specimens, we thank: J. Vindum and R. Drewes (CAS); A. Resetar (FMNH); R. Feeney (LACM); C. Austin (LSUMNS); S. Sweet (MSE); J. McGuire and C. Conroy (MVZ); A. Wynn (NMNH); P. Collins (SBMNH); B. Hollingsworth (SDMNH); C. Bell and R. Burroughs (TMM); T. LaDuc and R. Burroughs (TNHC); P. Holroyd (UCMP); R. Symonds (UMZC); J. Buskirk. We are greatful to S. Sweet for field assistance and the California Department of Fish and Game for permits. Much of the data collection was funded by NSD DBI-0306158 (to KDA).

@@

% figures
\begin{figure}[ht]
  \centering
  \includegraphics{figure/plastra}
  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
  \label{fig:plastra}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width = \textwidth]{figure/pca_imp}
  \caption{Pairs plot of the first three most important variables of the optimal random forest model of turtle plastral shape. The variables descend in importance from the upper left to the lower right. The observations are colored as in figure \ref{fig:gen_map}.  The bottom row are histograms of classification occurrences along the PCs.}
  \label{fig:imp_pc}
\end{figure}

\pagebreak
% tables
\begin{table}
  \centering
  \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
    \cline{3-4} 
    & & \multicolumn{2}{ c |}{Predicted class} \\ \cline{3-4}
    & & 1 & 0 \\ \hline
    \multicolumn{1}{| c |}{\multirow{2}{*}{Actual class}}
    & 1 & TRUE \newline POSITIVE & FALSE \newline NEGATIVE \\ \cline{2-4}
    \multicolumn{1}{| c |}{} & 0 & FALSE \newline POSITIVE & TRUE \newline NEGATIVE \\
    \hline
  \end{tabular}
  \caption{Example confusion matrix. The columns correspond to the predicted class of an observation, while the rows correspond to the actual class of that observation. Depending on the type match between the prediction and reality, four different outcomes are possible: true positive (TP), false negative (FN), false positive (FP), and true negative (TN). These four quantities are used for calculating all confusion matrix statistics. Each of these values is an integer and the sum of the number of occurrences of that event during classification.}
  \label{tab:conf}
\end{table}
